---
author: "Matthew Arsenault and Kelvin Bacelli"
title: "Effects on Cost of Health Insurance"
date: "2025-12-12"
output:
  pdf_document:
    number_sections: true
---
```{r echo=FALSE, include=FALSE, message=FALSE}
# all library load statements here

r.version <- "4.3.2"
seed.val <- 123456
```

Note: Brief intro and your main question go in this space. You may edit the headings and outline as fits your analysis. 

# Data Description and Processing

This data set is frequently taken from publicly accessible health insurance expense information on individuals with health insurance.

This data set contains descriptions of the individuals as well as how much they are charged for their insurance.\
Outcomes: \
charges: The medical insurance cost that the individual get charged. \
Descriptors:\
age: The age of the individual.\
sex: The sex of the individual.\
bmi: The body mass index of the individual that measures if they are at a health weight when considering height.\
children: The amount of children that the individual has.\
smoker: Either true or false of if the individual is a smoker.\
region: The area that the individual is located in.

This data set was checked for missing values. Of the original 1338 rows, only 3 of them contianed missing values. After the removal of those rows, we are left with 1335 for examination.

```{r echo=FALSE, include=FALSE}
data.df <- read.csv("insurance.csv")
data.df <- na.omit(data.df)
data.df$smoker <- factor(data.df$smoker)
data.df$region <- factor(data.df$region)
```


# Data Exploration


# Modeling 


## Hypothesis


## Model 1:


## Model 2: 


# Predicting with Models 1 and 2


# Summary and Conclusions

